---
title: "Pooled testing"
author: "Alexander V. Alekseyenko"
date: "5/26/2020"
output:
  pdf_document: default
  html_document: default
---


Suppose we have to screen N subjects with K testing capacity (K<=N). Also suppose for each subject 1,..., N we have (prior) estimates of them being positive, p_1, ..., p_N. Likewise, assume that up to 30 subjects can be pooled together without loss of sensitivity. See: https://www.thelancet.com/journals/laninf/article/PIIS1473-3099(20)30362-5/fulltext.

Probability that a pool of M subjects contains at least one positive is $\mathcal{P}_M(p_1,\ldots,p_M) = 1-\prod_i=1^M (1-p_i)$. Note that $1-(1-min(p_1,\ldots, p_M))^M \leq \mathcal{P}_M(p_1,\ldots, p_M) \leq 1-(1-max(p_1,\ldots, p_M))^M$.

```{r}
pool_positive = function(probs){
  1 - prod(1-probs)
}
```

Assuming that each sample in a pool is to be re-tested if the entire pool tests positive. 
The expected number of unit tests to screen M (M>1) samples is 
$\mathcal{E}_M(p_1,\ldots,p_M) = \\ (1-\mathcal{P}_M(p_1,\ldots,p_M)) + \mathcal{P}_M(p_1,\ldots,p_M)(1+M) = \\ 1+M\mathcal{P}_M(p_1,\ldots,p_M)$.

```{r}
expected_tests = function(probs){
  1+pool_positive(probs)*length(probs)
}
```


Note other strategies a possible here. For example, if a pool tests positive it can be sub-pooled into several smaller size pools. This may provide additional efficiency, but may have larger time requirements, so this is not considered here right now.

Example:

Compute the probabilities and the number of tests.
```{r}
ps = seq(from=0.005, to=0.27, by=0.01)
ns = 2:20
res= c()
for(p in ps){
  res = rbind(res,
              c(p, 1, p, 1))
  for(n in ns){
    res = rbind(res,
                c(p, n, pool_positive(rep(p,n)), expected_tests(rep(p,n))))
  }
}
res = as.data.frame(res)
colnames(res) = c("prob_positive", "pool_size", "pool_positive", "expected_tests")
```

Capacity gain is defined as the ratio between the samples needed to be tested and the expected number of tests using the pooling strategy.

```{r}
res$best = F
res$capacity_gain = res$pool_size/res$expected_tests
for(p in ps){
  res[res$prob_positive ==p,]$best = 
    (res[res$prob_positive ==p, ]$capacity_gain == max(subset(res, prob_positive==p)$capacity_gain))
}
```

```{r}
library(ggplot2)
library(reshape2)
```


```{r}
head(res)
```


```{r}
cgplot = 
ggplot(res, aes(y=capacity_gain, x = pool_size, group=prob_positive, color = prob_positive)) +
  theme_minimal() + 
  geom_line() + 
  geom_hline(yintercept=1, lty="dotted") + 
  geom_point(aes(size=best)) +
  scale_color_gradient(name="Positive rate", 
                       breaks = ps[seq(1,27,by=5)],
                       low = "black", high="grey75") +
  scale_size_discrete(guide="none") +
  ylab("Expected capacity gain") +
  xlab("Pool size")
print(cgplot)
```

```{r}
pdf("../results/cgPool.pdf", width=6, height=5)
print(cgplot)
dev.off()
```


```{r results='asis'}
library(knitr)
kable(subset(res, best))
```


```{r}
best_pool_matrix = 
  with(res, 
       tapply(capacity_gain, 
              list(prob_positive, pool_size), 
              max))

bps = data.frame(capacity_gain=apply(best_pool_matrix, 1, max),
      best_pool_size=as.numeric(colnames(best_pool_matrix)[apply(best_pool_matrix, 1, which.max)]))
bps$positive_rate = as.numeric(rownames(bps))
data.frame(pool_size = with(bps, tapply(best_pool_size, best_pool_size, min)),
           min_positive = with(bps, tapply(positive_rate, best_pool_size, min)),
           max_positive = with(bps, tapply(positive_rate, best_pool_size, max)),
           min_capacity_gain = with(bps, tapply(capacity_gain, best_pool_size, min)),
           max_capacity_gain = with(bps, tapply(capacity_gain, best_pool_size, max)))
```

```{r}
gg = ggplot(melt(best_pool_matrix), 
       aes(fill=value, x=Var1, y=Var2)) + 
  geom_tile() + 
  scale_fill_gradient2(low="red", mid="white", high="green", midpoint=1) +
  geom_text(aes(label=format(value, digits = 2)), size=1.2) +
  theme_minimal() + 
  ylab("Pool size")+xlab("Positive rate") + theme(legend.position = "none")
pdf("../results/posRateVSpoolSize.pdf", width=6, height=5)
print(gg)
dev.off()
print(gg)
```


# Streaming pooling algorithm

Suppose the current pool is of size $i$ and the probability that the pool is poisitve is $p_i$. A new specimen with probability of being positive equal to $p$ is proposed to be added to the pool. The probability that the pool containing these $i+1$ is positive is then $P(p_i, p) = p_i + (1-p_i)*p$. The expected number of tests for the $i+1$ specimens is then $P(p_i, p)(1+i+1) + (1-P(p_i, p)) = 1 + P(p_i, p)(1+i) = 1 + (p_i+(1-p_i)p)(i+1)$

```{r}
# Capacity gain for a pool test of n specimens with *pool* positive prob pp
cnp = function(n, pp){
  n/(1 + n*pp)
}

# Probability of pool being positive by adding a new specimen with individual probability pi
# to a pool with *pool* positive probability pp
add1pool_positive = function(pi, pp){
  pp + (1-pp)*pi
}

# Capacity gain by additing a new specimen with individual positvie probability pi 
# to a pool of n-1 specimens with *pool* positive probability pp
cnnp = function(pi, pp, n){
  n / (1 + n * add1pool_positive(pi, pp))
}
```


```{r}
pi = 0.05
p = 0.05
for(i in 2:10){
  if(cnp(i-1, p) > cnnp(pi, p, i)){
    print(i-1)
    break
  }
  p = add1pool_positive(pi, p)
}
p
pool_positive(rep(pi, 5))
c(4, cnp(4, pool_positive(rep(pi, 4))))
c(5, cnp(5, pool_positive(rep(pi, 5))))
c(6, cnp(6, pool_positive(rep(pi, 6))))
c(7, cnp(7, pool_positive(rep(pi, 7))))
```







